Visualizing probability distributions across bivariate nonlinear temporal granularities

### 2020-02-13 main.Rmd
Questions regarding section 2

\begin{definition}[Groups quasi-periodically]\label{def:quasi}
A granularity $H$ is quasi-periodical with respect to a granularity $G$ if:
(1) $G \trianglelefteq H$; and
(2) there exist a set of intervals $E_1,\dots,E_z$ (the granularity exceptions) and positive integers $R$, $P$, where $R$ is less than the minimum of the number of granules of $H$ between any two exceptions, such that for all $i \in Z$, if $H(i) = \bigcup_{k \in [0,k]} G(j_r)$  and $H (i + R) \neq \phi$ and $i + R < min(E)$, where E is the closest existing exception after $H (i)^2$, then $H (i + R) = \bigcup_{k \in [0,k]} G(j_r + P )$.
\end{definition}

@@

@@

@@

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra [@Ning2002-tf]. It is assumed that there exists a “bottom” granularity and calendar algebra operations are designed to generate
new granularities from the bottom one or recursively, from those already generated.
The calendar algebra consists of two kinds of operations: grouping-oriented and
granule-oriented operations. The grouping-oriented operations combine certain granules of a granularity together to form the granules of the new granularity. Example can be to consider a calendar with only two linear granularities "minute" and "hour" and "hour" is generated by grouping every $60$ "minutes". The granule-oriented operations do not change the granules of a granularity, but rather make choices of which granules should remain in the new granularity. For example, one can choose to look at the granularity "Monday" and hence select only "Monday" from linear granularity "day".

Some relevant grouping oriented operations are discussed, which will be used in Section \ref{sec:cyclic-gran} to define circular and aperiodic granularities.

- **The grouping operation** : Let $G1$ be a full-integer labeled granularity, and $m$ a positive integer. The grouping operation $Group_m(G)$ generates a new granularity $G2$, by partitioning the granules of $G$ into m-granule groups and making each group a granule of the resulting granularity.
More precisely, $G2 = Group_m(G1)$ is the full-integer labeled granularity such that for
each integer $i$, $G2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G1(j)$.

**Examples** :

  - $minute = Group_{60}(second)$
  - $hour = Group_{60}(minute)$,
  - $day = Group_{24}(hour)$,
  - $week = Group_{7}(day)$,

where, second(1) would start a minute and likewise.

- **The altering-tick operation** : Let G1, G2 be full-integer labeled granularities, and $l$, $k$, $m$ integers, where G2 partitions G1, and $1 \leq l \leq m$. The altering-tick operation $Alter^{m}_{l,k} (G2, G1)$ generates a new full-integer labeled granularity by periodically expanding or shrinking granules of G1 in terms of granules of G2. The altering-tick operation modifies the granules of G1 so that the $l$th granule of each group has $|k|$ additional (or fewer when k < 0) granules of G2.

**Example** :

\begin{equation*}
\begin{split}
pseudomonth = Alter^{12}_{11,-1}(day, Alter^{12}_{9,-1}(day, \\
  Alter^{12}_{6,-1}(day, Alter^{12}_{4,-1)}(day, \\
  Alter^{12}_{(2,-3)}(day, Group_{31}(day))))))
\end{split}
\end{equation*}

, where the granularity pseudomonth is generated by grouping 31 days, and then shrink April (4), June (6), September (9) and November (11) by 1 day, and shrink February (2) by 3 days. [@Ning2002-tf]

For more variations of grouping operations and granule oriented operations, the readers are recommended to see @Ning2002-tf.

# Cyclic time granularities {#sec:cyclic-gran}

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and key is a set of variables that define observational units over time. In a tsibble each observation should be uniquely identified by index and key.

A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble. Lining up with the the assumption in @Bettini2000-vy, that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists. In a cyclic organization of time, the domain is composed of a set of recurring time values, where any time value can be preceded and succeeded by another time value (for e.g Monday comes before Wednesday but Monday also succeeds Wednesday). Hence intuitively, cyclic granularities are additional abstractions of time which are not linear and hence the index order of linear granularities needs to be manipulated so that it leads to repetitive categorization of time.

Cyclic time granularities which accommodate for periodicities can be constructed by relating two linear granularities. The mappings between these linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might have days ranging from 28 to 31. To elaborate more, mappings will be regular if, for example, linear granularities are formed by grouping operation. Grouping operations combine fixed granules of a linear granularity to form a new granule of the resulting linear granularity. However, if granularities are formed by the altering-tick operation, for example, granules of the resulting granularity is composed of different number of granules of the root granularity. Cyclic time granularities are referred to as **circular** if mappings between linear time granularities are regular and **quasi-circular** if the same is irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month. The formalism would differ for circular and quasi-circular granularities although both accommodate for periodicities in data.

@@

Aperiodic cyclic granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$, that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity. Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to \autoref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that
\begin{equation*}
M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)
\end{equation*}
and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{th}$ granule of the $i^{th}$ linear aperiodic granularity.

\begin{definition}\label{aperiodic}
(aperiodic cyclic granularity): An aperiodic cyclic granularity $A_{B, M}$ relates aperiodic linear granularity $M$ and bottom granularity $B$, if

\begin{equation}\label{eq:aperiodic}
\begin{split}
A_{B, M}(z) & = L(i), \quad z \in T_{i_j} \quad and \quad \forall i \in {1, 2, \dots, n}\\
     & = L(0), \quad otherwise
\end{split}
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$T_{i_j}$ are the sets of indices of $B$ describing aperiodic linear granularities $M_{i}(j)$ such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$, $M = \bigcup_{i=1}^{n}M_{i}$.
\end{definition}

@@

```{r aperiodic-example, echo=FALSE,out.width = "100%",fig.cap="Aperiodic cyclic granularity $A(B, M1)$ relating aperiodic linear granularity $M1$ and bottom granularity $B$"}
knitr::include_graphics("Figs/aperiodic-example.png")
```

@@

In \autoref{sec:cyclic-gran}, we discussed how we can define cyclic granularities by extending the framework of linear granularities defined in @Bettini1998-ed. In this section, we will see how to obtain cyclic granularities through an algebraic representation of other cyclic granularities. This is similar to calendar algebra in @Ning2002-tf, where linear granularities are generated from the bottom linear granularity or from those that are already generated. Since, our method caters to computation of cyclic granularities, we shall refer to it as "cyclic calendar algebra".

>>>>

\todo[inline]{Give an example}

\begin{definition}[Groups quasi-periodically]\label{def:quasi}
A granularity $H$ is quasi-periodical with respect to a granularity $G$ if:
(1) $G \trianglelefteq H$; and
(2) there exist a set of intervals $E_1,\dots,E_z$ (the granularity exceptions) and positive integers $R$, $P$, where $R$ is less than the minimum of the number of granules of $H$ between any two exceptions, such that for all $i \in Z$, if $H(i) = \bigcup_{k \in [0,k]} G(j_r)$  and $H (i + R) \neq \phi$ and $i + R < min(E)$, where $E$ is the closest existing exception after $H (i)^2$, then $H (i + R) = \bigcup_{k \in [0,k]} G(j_r + P)$.
\end{definition}

@@

\todo[inline]{Give an example}

@@

\todo[inline]{What is g-rel?}

@@

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra [@Ning2002-tf]. It is assumed that there exists a “bottom” granularity and calendar algebra operations are designed to generate new granularities recursively from the bottom up. The calendar algebra consists of two kinds of operations: grouping-oriented and granule-oriented operations. The grouping-oriented operations combine certain granules of a granularity together to form the granules of the new granularity. For example, consider a calendar with only two linear granularities: "minute" and "hour", where "hour" is generated by grouping every $60$ "minutes". The granule-oriented operations do not change the granules of a granularity, but rather make choices of which granules should remain in the new granularity. For example, one can choose to look at the granularity "Monday" and hence select only "Monday" from linear granularity "day".

Some relevant grouping-oriented operations are discussed below; these will be used in Section \ref{sec:cyclic-gran} to define circular and aperiodic granularities.

- **The grouping operation**: Let $G_1$ be a full-integer labeled granularity, and $m$ a positive integer. The grouping operation Group$_m(G)$ generates a new granularity $G_2$, by partitioning the granules of $G_1$ into $m$-granule groups and making each group a granule of the resulting granularity. More precisely, $G_2 = \text{Group}_m(G_1)$ is the full-integer labeled granularity such that for each integer $i$,
$$G_2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G_1(j).$$

\todo[inline]{What does "full-integer" mean?}

For example:
\begin{align*}
  \text{minute} &= \text{Group}_{60}(\text{second});\\
  \text{hour}   &= \text{Group}_{60}(\text{minute});\\
  \text{day}    &= \text{Group}_{24}(\text{hour});\\
  \text{week}   &= \text{Group}_{7}(\text{day});
\end{align*}

where $\text{second}(1)$ would start a minute and likewise.

\todo[inline]{What does this last phrase mean?}

- **The altering-tick operation** : Let $G_1$, $G_2$ be full-integer labeled granularities, and $l$, $k$, $m$ integers, where $G_2$ partitions $G_1$, and $1 \leq l \leq m$. The altering-tick operation $\text{Alter}^{m}_{l,k} (G_2, G_1)$ generates a new full-integer labeled granularity by periodically expanding or shrinking granules of $G_1$ in terms of granules of $G_2$. The altering-tick operation modifies the granules of $G_1$ so that the $l$th granule of each group has $|k|$ additional (or fewer when $k < 0$) granules of $G_2$.

For example, due to the uneven length of months, we need to use the altering-tick operation to compute months from days.
\begin{align*}
\text{pseudomonth} = &\text{Alter}^{12}_{11,-1}(\text{day}, \\
  &~~\text{Alter}^{12}_{9,-1}(\text{day}, \\
  &~~~~\text{Alter}^{12}_{6,-1}(\text{day}, \\
  &~~~~~~\text{Alter}^{12}_{4,-1)}(\text{day}, \\
  &~~~~~~~~\text{Alter}^{12}_{(2,-3)}(\text{day}, \\
  &~~~~~~~~~~\text{Group}_{31}(\text{day})))))),
\end{align*}
where the granularity pseudomonth is generated by grouping 31 days, and then shrink April (4), June (6), September (9) and November (11) by 1 day, and shrink February (2) by 3 days [@Ning2002-tf].

\todo[inline]{Why is this called a ``pseudomonth''?}

For more variations of grouping operations and granule oriented operations, see @Ning2002-tf.

# Nonlinear time granularities {#sec:cyclic-gran}

@@

We propose a formalism of nonlinear time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and key is a set of variables that define observational units over time. In a tsibble each observation should be uniquely identified by index and key.

A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble. Lining up with the the assumption in @Bettini2000-vy, that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists. In a cyclic organization of time, the domain is composed of a set of recurring time values, where any time value can be preceded and succeeded by another time value (for e.g Monday comes before Wednesday but Monday also succeeds Wednesday). Hence intuitively, nonlinear granularities are additional abstractions of time which are not linear and hence the index order of linear granularities needs to be manipulated so that it leads to repetitive categorization of time.

Nonlinear time granularities which accommodate for periodicities can be constructed by relating two linear granularities. The mappings between these linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might have days ranging from 28 to 31. To elaborate more, mappings will be regular if, for example, linear granularities are formed by grouping operation. Grouping operations combine fixed granules of a linear granularity to form a new granule of the resulting linear granularity. However, if granularities are formed by the altering-tick operation, for example, granules of the resulting granularity is composed of different number of granules of the root granularity. Nonlinear time granularities are referred to as **circular** if mappings between linear time granularities are regular and **quasi-circular** if the same is irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month. The formalism would differ for circular and quasi-circular granularities although both accommodate for periodicities in data.

@@

Aperiodic nonlinear granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$, that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity. Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to \autoref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that
\begin{equation*}
M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)
\end{equation*}
and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{th}$ granule of the $i^{th}$ linear aperiodic granularity.

\begin{definition}[Aperiodic nonlinear granularity]\label{aperiodic}
An aperiodic nonlinear granularity $A_{B, M}$ relates aperiodic linear granularity $M$ and bottom granularity $B$, if
\begin{equation}\label{eq:aperiodic}
\begin{split}
A_{B, M}(z) & = L(i), \quad z \in T_{i_j} \quad and \quad \forall i \in {1, 2, \dots, n}\\
     & = L(0), \quad otherwise
\end{split}
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$T_{i_j}$ are the sets of indices of $B$ describing aperiodic linear granularities $M_{i}(j)$ such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$, $M = \bigcup_{i=1}^{n}M_{i}$.
\end{definition}

@@

```{r aperiodic-example, echo=FALSE,out.width = "100%",fig.cap="Aperiodic nonlinear granularity $A(B, M1)$ relating aperiodic linear granularity $M1$ and bottom granularity $B$"}
knitr::include_graphics("Figs/aperiodic-example.png")
```

@@

In \autoref{sec:cyclic-gran}, we discussed how we can define nonlinear granularities by extending the framework of linear granularities defined in @Bettini1998-ed. In this section, we will see how to obtain cyclic granularities through an algebraic representation of other cyclic granularities. This is similar to calendar algebra in @Ning2002-tf, where linear granularities are generated from the bottom linear granularity or from those that are already generated. Since, our method caters to computation of cyclic granularities, we shall refer to it as "cyclic calendar algebra".
